Properties of winning Iterated Prisoner's Dilemma strategies.

### 2020-04-13 paper/main.tex

A result summary (Table~\ref{table:output_result}) has \(N\) number of rows
because each row contains information for each strategy that participated in the
tournament. The information includes the strategy's rank, median score, the rate
with which the strategy cooperated $(C_r)$, its match win count, and the
probability that the strategy cooperated in the opening move. Moreover, the
probabilities of a strategy being in any of the four states ($CC, CD, DC, DD$),
and the rate of which the strategy cooperated after each state. The \textbf{normalised rank}
feature that is manually added. The rank \(R\) of a given
strategy can vary between 0 and \(N-1\). Thus, the normalised rank,
denoted as $r$, is calculated as a strategy's rank divided by \(N - 1\).

@@

    algebraic distance between them, defined as SSE. A SSE value of 1 indicates

@@

    strategy is behaving a ZDs. The rest of the features considered are the $CC$

@@

The memory usage of strategies with an infinite memory size, for example Evolved
FSM 16 Noise 05, is equal to 1. Otherwise the memory usage is the number of
rounds of play used by the strategy divided by the number of turns in each match.
For example, Winner12 uses the previous two rounds of play, and if participating in a
participated in a tournament where $n$ was 100 the memory usage would be 2/100.
Note that for tournaments with a probabilistic
ending the number of turns was not collected, so the memory usage feature is not
used for probabilistic ending tournaments.

@@

Moreover, the manner in which a strategy achieves a given cooperation rate
relative to the tournament population average is important. Playing a strategy
that randomly cooperates with $C_{\text{mean}}$ is unlikely to be as effective as it will
randomly have cooperations matched with defections. In contrast, TFT naturally
achieves a cooperation rate near $C_{\text{mean}}$ by virtue of copying its opponent's
last move while also minimizing instances where it is exploited by an opponent
(cooperating while the opponent defects), at least in non-noisy tournaments.
\footnote{This also explains why Tit For N Tats does not fare well -- it fails
to achieve the proper cooperation ratio.}
TFT forces the opponent to pay back the (C, D) round with a (D, C) round before
returning to mutual cooperation. This explains why TFT performed well
in Axelrod's original tournaments as most strategies submitted to those tournaments
were typically cooperative and relatively few strategies used an easily exploitable
pattern. TFT does not appear in the top ranks in
these tournaments because it is too nice. Strategies like Grudger will always
defect after a fixed number of opponent defections, which allows them to
effectively exploit strategies like Alternator or stochastic strategies that
have a non-zero chance of cooperating after mutual defection, which TFT will
not do. Moreover in a noisy environment these strategies will naturally tend
toward always defecting, leading them to exploit strategies like Cooperator. In
such a scenario the noisy environment effectively voids Axelrod's rule to be nice,
allowing strategies to attempt exploitation, whereas in a noise-free environment,
exploitation is risky because several strategies exhibit a Grudger-like behavior,
reducing the overall value in attempting to exploit strategies like Cooperator.

Similarly, these results suggest an explanation regarding the intuitively unexpected
effectiveness of memory one strategies historically. Given that among the important
features associated with success are the relative cooperation rate to the population
average and the four memory-one probabilities of cooperating conditional on the
previous round of play, all five features can be optimized by a memory one
strategy such as TFT. Usage of more history becomes valuable when there are
exploitable opponent patterns, indicated by the importance of SSE as a feature,
that the first-approximation provided by a memory one strategy is no longer
sufficient.

@@

A multivariate linear regression has been fitted to model the relationship between
the features and the normalised rank. Based on the graphical representation of
the correlation matrices given in Appendix~\ref{app:correlations} several of the
features are highly correlated. Highly correlated features have been removed
before fitting the linear regression model. The features included are given
by Table~\ref{table:linear_regression} alongside their corresponding \(p\) values
in the distinct tournaments and their regression coefficients.

@@

A multivariate linear regression has also be fitted on the median score. The
coefficients and \(p\) values of the features can be found in
Appendix~\ref{app:regression_median_score}. The results of the two methods
are in agreement.

@@

A third method that evaluates the importance of the features in
Table~\ref{table:manual_features} using clustering and random forests can be found in
the Appendix~\ref{app:clustering}. The results uphold the outcomes of the
correlation and multivariate regression. It also evaluates the effects
of the classifiers stochastic, make use of game, and make use of length which
have not been evaluated by the methods above because there are binary variables.
The results imply that they have no significant effect on a strategy's
performance.

@@

So what is the best way of playing the IPD? And is there a single dominant
strategy for the IPD? There was not a single strategy within the collection of the
\numberofstrategies strategies, that has managed to perform well in all the
tournaments variations it competed in.
Even if on average a strategy ranked highly in a specific environment it did not
guarantee its success over the different tournament types. However, the results
of sections~\ref{section:top_performances}
and~\ref{section:evaluation_of_performance} have demonstrated that there are
properties associated with the success of strategies. A few of the properties that
have been identified by this manuscript's analysis contradict the properties of
Axelrod~\cite{Axelrod1981}. Namely, in Section~\ref{section:top_performances} it
was shown that trained strategies and strategies that decided their actions
based on pre-designed strategies to maximise their utility dominated several
tournaments across tournament types, hinting that successful IPD strategies are
often clever or more complex than simple strategies like TFT. Most of the
successful strategies highlighted
in~Section~\ref{section:top_performances} were strategies that begin with
cooperation.

@@

Thus, the five properties successful strategies need to have in a IPD competition
are: be nice, be provocable and contrite, be a little envious, be clever, and adapt to the
environment (including the population of strategies).

>>>>

A result summary (Table~\ref{table:output_result}) has \(N\) number of rows
because each row contains information for each strategy that participated in the
tournament. The information includes the strategy's rank, median score, the rate
with which the strategy cooperated $(C_r)$, its match win count, and the
probability that the strategy cooperated in the opening move. Moreover, the
probabilities of a strategy being in any of the four states ($CC, CD, DC, DD$),
and the rate of which the strategy cooperated after each state. The \textbf{normalised rank}
feature that is manually added. The rank \(R\) of a given. % TODO Clarify how strategies are ranked (see email discussion)
strategy can vary between 0 and \(N-1\). Thus, the normalised rank,
denoted as $r$, is calculated as a strategy's rank divided by \(N - 1\).

@@

    algebraic distance between them as the sum of squared error (SSE). A SSE value of 1 indicates

@@

    strategy is behaving as a ZDs. The rest of the features considered are the $CC$

@@

The memory usage of strategies is the number of
rounds of play used by the strategy divided by the number of turns in each match. 
For example, Winner12 uses the previous two rounds of play, and if participating in a
participated in a match with 100 turns, the memory usage would be 2/100.
For strategies with an infinite memory size, for example Evolved
FSM 16 Noise 05, is equal to 1.
Note that for tournaments with a probabilistic
ending the number of turns was not collected, so the memory usage feature is not
used for probabilistic ending tournaments.

@@

Moreover, the manner in which a strategy achieves a given cooperation rate
relative to the tournament population average is important. Playing a strategy
that randomly cooperates with $C_{\text{mean}}$ is unlikely to be as effective as it will
randomly have cooperations matched with defections. In contrast, TFT naturally
achieves a cooperation rate near $C_{\text{mean}}$ by virtue of copying its opponent's
last move while also minimizing instances where it is exploited by an opponent
(cooperating while the opponent defects), at least in non-noisy tournaments.
\footnote{This also explains why Tit For N Tats does not fare well -- it fails
to achieve the proper cooperation ratio.}
TFT forces the opponent to pay back the (C, D) round with a (D, C) round before
returning to mutual cooperation. This explains why TFT performed well
in Axelrod's original tournaments as most strategies submitted to those tournaments
were typically cooperative and relatively few strategies used an easily exploitable
pattern. TFT does not appear in the top ranks in
 tournaments with more aggressive strategies because it is too nice. Strategies like Grudger will always
defect after a fixed number of opponent defections, which allows them to
effectively exploit strategies like Alternator or stochastic strategies that
have a non-zero chance of cooperating after mutual defection, which TFT will
not do. Moreover in a noisy environment these strategies will naturally tend
toward always defecting, leading them to exploit strategies like Cooperator. In
such a scenario the noisy environment effectively voids Axelrod's rule to be nice,
allowing strategies to attempt exploitation, whereas in a noise-free environment,
exploitation is risky because several strategies exhibit a Grudger-like behavior,
reducing the overall value in attempting to exploit strategies like Cooperator.

Similarly, these results suggest an explanation regarding the intuitively unexpected
effectiveness of memory one strategies historically. Given that among the important
features associated with success are the relative cooperation rate to the population
average and the four memory-one probabilities of cooperating conditional on the
previous round of play, all five features can be optimized by a memory one
strategy such as TFT. Usage of more history becomes valuable when there are
exploitable opponent patterns. This is indicated by the importance of SSE as a feature,
showing that the first-approximation provided by a memory one strategy is no longer
sufficient.

@@

A multivariate linear regression has been fitted to model the relationship between
the features and the normalised rank. Based on the graphical representation of
the correlation matrices given in Appendix~\ref{app:correlations} several of the
features are highly correlated and have been removed
before fitting the linear regression model. The features included are given
by Table~\ref{table:linear_regression} alongside their corresponding \(p\) values
in the distinct tournaments and their regression coefficients.

@@

A multivariate linear regression has also be fitted on the median score. The
coefficients and \(p\) values of the features can be found in
Appendix~\ref{app:regression_median_score}. This approach leads to similar conclusions.

@@

A third method that evaluates the importance of the features in
Table~\ref{table:manual_features} using clustering and random forests can be found in
the Appendix~\ref{app:clustering}. The results uphold the outcomes of the
correlation and multivariate regression. It also evaluates the effects
of the whether or not a strategy is stochastic, makes use of the knowledge of the utility values, or makes use of match length. These were not evaluated by the methods above because there are binary variables.
The results showed that they have no significant effect on a strategy's
performance.

@@

So what is the best way of playing the IPD? And is there a single dominant
strategy for the IPD?

There was not a single strategy within the collection of the
\numberofstrategies strategies, that has managed to perform well in all the
tournaments variations it competed in.
Even if on average a strategy ranked highly in a specific environment it did not
guarantee its success over the different tournament types. However, the results
of sections~\ref{section:top_performances}
and~\ref{section:evaluation_of_performance} have demonstrated that there are
properties associated with the success of strategies. A few of the properties that
have been identified by this manuscript's analysis contradict the properties of
Axelrod~\cite{Axelrod1981}. Namely, in Section~\ref{section:top_performances} it
was shown that trained strategies dominated several
tournaments across tournament types, hinting that successful IPD strategies are
often clever or more complex than simple strategies like TFT. Most of the
successful strategies highlighted
in~Section~\ref{section:top_performances} were strategies that begin with
cooperation.

@@

Thus, the five properties successful strategies need to have in a IPD competition
are:

\begin{itemize}
\item be nice, 
    \item be provocable and contrite, 
    \item be a little envious, 
    \item be clever, 
    \item adapt to theenvironment (including the population of strategies).
\end{itemize}
